Reasonably Programmable Syntax

### 2017-06-06 popl18/syntax-popl18.tex
switch to acmsmall template, collapsing abstract, reorganized intro

\PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}
%% For double-blind review submission
\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission
%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission
%\documentclass[acmlarge]{acmart}\settopmatter{}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
%Many languages define convenient list literal forms. This paper questions this common practice: why should an otherwise ordinary data structure be privileged in the definition of a general-purpose programming language? 
When designing the surface syntax of a programming language, it is common practice to build in literal forms for a select few standard data structures, e.g. lists. The problem is that this practice neglects an unbounded assortment of other data structures for which literal forms might also be useful (consider, for example, sets, maps, matrices, regular expressions, HTML trees, URLs, SQL queries and chemical structures.) A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner. However, the mechanisms available to programmers today, e.g. Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are \emph{unreasonable}. In particular, they do not support modular reasoning about syntactic determinism, so separately defined literal forms can and do conflict syntactically with one another. Moreover, clients cannot reason abstractly about types and binding when they encounter an unfamiliar literal form (like they can when they encounter an unfamiliar function being applied.) Instead, they must reason transparently, i.e. about the underlying expansion. This increases cognitive cost, defeating much of the purpose of convenient literal forms. %This makes it difficult to program ``in the large'', and these mechanisms have not been widely adopted.% Hygienic, typed term-rewriting systems, e.g. the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.

This paper formally details a more reasonable alternative: \emph{typed literal macros} (TLMs). TLMs give library providers full programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} \cite{TSLs} at a specified type or parameterized family of types. Expansion is {strictly hygienic}, meaning that 1) expansions cannot make any assumptions about the application site context; and 2)  expansions do not reveal internal bindings to spliced terms or to the remainder of the program. 
% Partial parameter application lowers the syntactic cost of this strict style. 
This design ensures that clients are able to reason modularly about syntactic determinism and abstractly about the type and binding structure of the program. The system needs only to convey to clients (e.g. via secondary notation) the \emph{segmentation} of each literal, which gives the locations of the spliced segments. 
%The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.) 
% This mechanism captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles. 
We establish these abstract reasoning principles formally with a calculus of expressions, patterns and ML-style modules.  This is the first detailed type-theoretic account of a typed, hygienic term-rewriting macro system for an ML-like\todo{or Scala-like?} language. % This calculus is the first such formalization of a typed macro system.

@@

\vspace{-4px}
\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
The surface syntax of a programming language serves as the interface between human programmers and the language's semantics. As such, language designers often include various \emph{literal forms} in the surface syntax that are designed to decrease the cognitive cost of constructing and pattern matching over values of certain types defined in the standard library. For example, in Standard ML (SML), the list expression literal \li{[x, y, z]} desugars to a term  semantically equivalent to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} identify the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the expansion of literal forms is \emph{context independent}. We will return to the concept of context independence throughout this work.} List pattern literals are analagous. Another example is Ur/Web, which extends Ur's textual syntax with expression and pattern literals for XHTML trees, as defined in Ur/Web's standard library \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}. Figure \ref{fig:urweb} shows examples of XHTML literals containing ``spliced'' string expressions (Line 1) and XHTML expressions (Line 2). 
%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.

Third-party library providers are quite justifiably envious of the special status of these few otherwise ordinary user-defined types. After all, it is not difficult to find other examples of types for which mathematicians, scientists and programmers have invented specialized notations. For example, 1) clients of a ``collections'' library might like not just list literals, but also vector, matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library might like not just XHTML literals but also CSS literals (which Ur/Web does not provide); 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}. \citet{Bravenboer:2007:PIA:1289971.1289975} suggests that literal forms for structured encodings of SQL queries and other query types can reduce the temptation to use string encodings and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}. Choosing strings over composite data structures, a.k.a. ``stringly-typed programming'', is quite common in practice -- \citet{TSLs} sampled strings from open source projects and found that at least 15\% of them were parseable according to some readily apparent type-specific grammar, e.g. for SQL queries, regexes, file system paths, URLs and others. Literal forms at these types, with support for splicing, would lower the cost of using more compositional encodings. %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.

\begin{figure}
\vspace{-2px}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-10px}
\caption{XHTML literals with support for splicing at two different types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
\vspace{-7px}
\label{fig:urweb}
\end{figure}

Language designers cannot, of course, hope to accomodate all such types with primitive literal forms. Instead, there has been considerable interest in mechanisms that allow programmers to introduce new literal forms in a decentralized manner. We will cover a number of existing mechanisms, including Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, in Sec. \ref{sec:existing-approaches}. The problem is that these existing mechanisms quickly become \emph{unreasonable} when ``programming in the large'' \cite{DeRemer76}.
%, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''
For example, consider the perspective of a programmer attempting to comprehend (i.e. reason about) the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of ML's (or Scala's) syntax where a large number of syntax extensions defining user-defined literal forms, some unfamiliar to the programmer, have been imported.

\begin{figure}[h]
\vspace{-10px}
\begin{lstlisting}
  val w = compute_w ()
  val x = compute_x ()
  val y = {|(!R)@&{&/x!/:2_!x}'!R}|}
\end{lstlisting}
\vspace{-10px}
\caption{An example of unreasonable program text.}
\vspace{-9px}
\label{fig:K-dialect}
\end{figure}

\noindent
If the programmer happens to be familiar with the intentionally terse syntax of the stack-based database query language K \cite{Whitney:2001:LOR:376284.375783}, then the literal form on Line 3 might be convenient. If the programmer does not recognize this syntax, however, there are no simple, definitive protocols for answering questions like:

\begin{enumerate}
\item \textbf{(Responsibility)} Which syntax extension determines the desugaring of the literal form on Line 3? Might there be many possible desugarings (i.e. syntactic non-determinism) due to conflicts between extensions?
\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 taken as spliced expressions (meaning that they appear directly in the desugaring), or are they parsed in some other way peculiar to this literal form?
\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2, or might it capture an unseen binding introduced in the desugaring of Line 3?
\item \textbf{(Context Dependence)} If \li{w}, on Line 1, is renamed, could that possibly break Line 3, or change its meaning? In other words, might the desugaring of Line 3 assume that some variable identified as \li{w} is in scope (even though \li{w} is not mentioned in the text of Line 3)?
\item \textbf{(Typing)} What type does \li{y} have?
\end{enumerate}

In short, the problem is that library providers cannot reason modularly about syntactic determinism (because other extensions might define overlapping forms), and if the desugaring of the program text is held abstract, programmers can no longer reason about types, binding and segmentation (i.e. answer questions like those above.) This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter a wide variety of libraries \cite{DBLP:conf/sac/LammelPS11}. Forcing the programmer to examine the desugaring of the program text, or the logic that produced that desugaring, in order to answer these sorts of basic questions about segmentation, binding and typing can defeat the ultimate purpose of syntactic sugar -- decreasing cognitive cost \cite{Green89}. Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala -- the binding discipline of the language ensures that programmers can reason lexically about binding and renaming, and types mediate abstraction over function and module implementations.  %From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.

This paper introduces a more reasonable approach: typed literal macros (TLMs). TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms}, which syntactically subsume other literal forms because their bodies are left initially unparsed according to the fixed context free syntax of the language (i.e. they are initially parsed like raw string literals.) The applied TLM parses the literal body and generates an expansion during a phase known as typed expansion, which generalizes the usual typing phase. For example, Fig. \ref{fig:K-tsm-example} applies a TLM to express the ``unreasonable'' example from Fig. \ref{fig:K-dialect}.

TLMs come equipped with uniquely powerful abstract reasoning principles. We will more precisely characterize these abstract reasoning principles as we proceed. For now, let us develop some informal intuitions -- the client programmer can reason about Line 3 as follows:
\begin{enumerate}
\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is solely responsible for typed expansion of the literal body. The context-free syntax of the language is fixed, so the TLM provider can modularly establish that the syntax the TLM implements is deterministic.
\item \textbf{(Segmentation)} The output that a TLM generates is structured so that the system can produce from it an accurate \emph{segmentation} of the literal body that distinguishes spliced segments, which appear directly in the expansion, from segments that are parsed in some other way by the TLM. This segmentation can be communicated by a pretty printer or code editor using secondary notation, e.g. colors in this document. By examining the segmentation of Line 3, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions, whereas \li{R} is parsed in some other way.
\item \textbf{(Capture)} The system avoids capture, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2. It cannot capture an unseen binding of \li{x} in the expansion of Line 3.
\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, for example, \li{w} is in scope. The client is therefore free to rename \li{w}.
\item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that the expansion must have. We will see an example of a TLM definition in the next section. Moreover, each spliced segment in the segmentation also comes paired with the type it is expected to have. This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
\end{enumerate}

@@

\begin{figure}
\vspace{-4px}
\begin{lstlisting}[numbers=none,xleftmargin=0px]
  val w = compute_w ()
  val x = compute_x ()
  val y = $kquery `SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL`
\end{lstlisting}
\vspace{-8px}
\caption{TLMs make examples like the one from Fig. \ref{fig:K-dialect} more reasonable.}
\vspace{-8px}
\label{fig:K-tsm-example}
\end{figure}

The remainder of this document is organized as follows.
\begin{itemize}
\item Sec. \ref{sec:setsms} begins by going through the mechanics of defining a {simple expression TLM} for XHTML and describes initial parsing, typed expansion and expansion validation in more detail.
\item Sec. \ref{sec:sptsms} does the same for a {simple pattern TLM}, i.e. a TLM that generates patterns. %The hygiene condition for pattern TLMs is interesting.
\item Sec. \ref{sec:setsms-formally} defines a calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above.

@@

% We next develop a more sophisticated account of TLMs in Chapters \ref{chap:ptsms} and \ref{chap:static-eval}, with the aim of working out details necessary to integrate TLMs into full-scale general-purpose languages like ML or Scala. %These two chapters constitute Part \ref{part:parametric-tsms} of our contributions.
\item Sec. \ref{sec:ptsms} adds parameterized types and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e. TLMs that take type and module parameters. Parameters serve two purposes:
\begin{enumerate}
\item They allow for TLMs that operate not just at a single type, but over a type- and module-parameterized family of types. For example, we consider a parametric TLM for dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the co-domain type.
\item Parameters also give expansions access to terms other than those explicitly spliced in. %This is particularly convenient for working with abstract types \emph{a la} ML. 
Partial parameter application in TLM abbreviations lowers the cost of this parameter-passing discipline.
\end{enumerate}
This section culminates with a calculus that extends the calculus of Sec. \ref{sec:setsms-formally} with ML-style modules and parametric TLMs.
\item Sec. \ref{sec:static-eval} considers the topic of term evaluation during the typed expansion phase, i.e. \emph{static evaluation}, in  more detail, and also gives examples of TLMs that are useful for defining other TLMs, e.g. TLMs that implement parser generators and quasiquotation.
\item Sec. \ref{sec:related-work} compares TLMs to the related work, including closely related work by \citet{TSLs,sac15} on type-specific languages. \item Sec. \ref{sec:discussion} concludes with a summary of our contributions and a brief discussion of present limitations and future research directions.
\end{itemize}

Due to space limitations, many standard or straightforward technical details of the calculi defined in Sec. \ref{sec:setsms-formally} through Sec. \ref{sec:static-eval} are relegated to the supplemental material. For convenience, the examples that we give are written in a full-scale functional language under development called VerseML. % For the present purposes, we can consider VerseML as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs. %VerseML is, as its name suggests, a conceptual descendent of ML. It diverges from other dialects of ML that have a similar type structure in that it has a bidirectional type system \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanism of TLM implicits described in Chapters \ref{chap:tsls} and \ref{chap:ptsms}. 
%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of VerseML in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TLMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work. 
% We will not formally define these features mainly to avoid unnecessarily complicating our presentation with details that are not essential to the ideas presented herein. As such, 
All examples written in VerseML should be understood to be informal motivating material for the subsequent formal material. %We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble. %The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification. 
%We will give a brief overview of these languages are organized in Sec. \ref{sec:VerseML}.

@@

\end{lstlisting}
\vspace{-5px}
\caption[Definitions of various types used by TLM definitions.]{Definitions of various types used by TLM definitions. We assume that these type definitions are available ambiently.}
\label{fig:indexrange-and-parseresult}
\label{fig:candidate-exp-verseml}
\vspace{-5px}
\end{figure}

>>>>

\PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}
%% For double-blind review submission
\documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission
%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission
%\documentclass[acmlarge]{acmart}\settopmatter{}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
%Many languages define convenient list literal forms. This paper questions this common practice: why should an otherwise ordinary data structure be privileged in the definition of a general-purpose programming language? 
When designing the surface syntax of a programming language, it is common practice to build in literal forms for a few standard data structures, e.g. lists. The problem is that this practice neglects an unbounded variety of other data structures for which literal forms might also be useful (consider, for example,  maps, regexes, HTML trees, URLs, SQL queries and chemical structures.) %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner. However, the mechanisms available to programmers today, e.g. Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are \emph{unreasonable}. In particular, they do not support modular reasoning about syntactic determinism, so separately defined literal forms can and do conflict syntactically with one another. Moreover, clients cannot reason abstractly about types and binding when they encounter an unfamiliar literal form (like they can when they encounter an unfamiliar function being applied.)% Instead, they must reason transparently, i.e. about the underlying expansion. This increases cognitive cost, defeating much of the purpose of convenient literal forms. %This makes it difficult to program ``in the large'', and these mechanisms have not been widely adopted.% Hygienic, typed term-rewriting systems, e.g. the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.
This paper introduces a system of \emph{typed literal macros (TLMs)} that allows library providers to define new literal forms in a decentralized manner. More specifically, TLMs give library providers full programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} \cite{TSLs} at a specified type or parameterized family of types. % The mechanism is {strictly hygienic}, meaning that 1) the expansion must be context-independent; and 2)  splicing is capture avoiding. 
% Partial parameter application lowers the syntactic cost of this strict style. 
Compared to existing approaches, including syntax extension systems and macro-based quotation rewriting systems, TLMs are uniquely \emph{reasonable}: TLM providers can reason modularly about syntactic determinism, and TLM clients can reason abstractly about the type and binding structure of the program. The system needs only to convey to clients (e.g. via secondary notation) the inferred \emph{segmentation} of each literal, which gives the locations of the spliced terms. The inferred segmentation also informs the hygiene mechanism (previously, only unhygienic macros could freely splice terms out of literal bodies.) 
%The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.) 
% This mechanism captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles. 
We establish these abstract reasoning principles formally with a calculus of expressions, patterns and ML-style modules. This is the  first detailed type-theoretic account of a typed, hygienic term-rewriting macro system for an ML-like language. % This calculus is the first such formalization of a typed macro system.

@@

% \vspace{-4px}
\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
The surface syntax of a programming language serves as the interface between human programmers and a semantics. As such, language designers often include in the surface syntax various \emph{literal forms}  that are designed to decrease the cognitive cost of constructing and pattern matching over values of certain otherwise ordinary types from the standard library. For example, in Standard ML (SML), the list expression literal \li{[x, y, z]} desugars to a term equivalent to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} identify the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the expansion of literal forms is \emph{context independent}. We will return to the concept of context independence again.} List pattern literals are analagous. Another example is Ur/Web, which extends Ur's textual syntax with expression and pattern literals for XHTML trees, as defined in Ur/Web's standard library \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}. Fig. \ref{fig:urweb} shows examples of XHTML literals containing ``spliced'' string expressions (Line 1) and XHTML expressions (Line 2). 
%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.

Third-party library providers are quite justifiably envious of the special status of these few otherwise ordinary standard library constructs. After all, it is not difficult to find other examples of types for which mathematicians, scientists and programmers have invented specialized notations. For example, 1) clients of a ``collections'' library might like not just list literals, but also matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library might like not just XHTML literals but also CSS literals (which Ur/Web does not provide); 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}. \citet{Bravenboer:2007:PIA:1289971.1289975} suggests that literal forms for structured encodings of SQL queries and other query types can reduce the temptation to use string encodings and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}. Choosing strings over composite data structures, a.k.a. ``stringly-typed programming'', is rather common in practice -- \citet{TSLs} sampled strings from open source projects and found that at least 15\% of them were parseable according to some readily apparent type-specific grammar, e.g. for SQL queries, regexes, file system paths, URLs and others. Literal forms at these types, with support for splicing, would lower the cost of compositional encodings.%Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.

\begin{figure}
\vspace{-2px}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-10px}
\caption{XHTML literals with support for splicing at two different types are built primitively into Ur/Web.}
\vspace{-7px}
\label{fig:urweb}
\end{figure}

Language designers cannot, of course, hope to accomodate all such types with primitive literal forms. Instead, there is considerable interest in mechanisms that allow programmers to define new literal forms in a decentralized manner. For example, we will consider a number of syntax extension mechanisms, including Camlp4 \cite{ocaml-manual} and SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, in Sec. \ref{sec:existing-approaches}. The problem is that these mechanisms quickly become \emph{unreasonable} when ``programming in the large'' \cite{DeRemer76}.
%, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''
For example, consider the perspective of a programmer attempting to comprehend (i.e. reason about) the program text in Fig. \ref{fig:K-dialect}, which is written in a dialect of ML's (or Scala's) syntax where a variety of syntax extensions defining user-defined literal forms, some unfamiliar to the programmer, are active.

\begin{figure}[h]
\vspace{-10px}
\begin{lstlisting}
  val w = compute_w()
  val x = compute_x(w)
  val y = {|(!R)@&{&/x!/:2_!x}'!R}|}
\end{lstlisting}
\vspace{-10px}
\caption{An example of unreasonable program text.}
\vspace{-9px}
\label{fig:K-dialect}
\end{figure}

\noindent
If the programmer happens to be familiar with the intentionally terse syntax of the stack-based database query language K \cite{Whitney:2001:LOR:376284.375783}, then the literal form on Line 3 might be convenient. If the programmer does not recognize this rather obscure syntax, however, there are no simple, definitive protocols for answering questions like:

\begin{enumerate}
\item \textbf{(Responsibility)} Which syntax extension determines the desugaring of the literal form on Line 3? Might there be conflicting desugarings (i.e. syntactic non-determinism)?
\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 taken as spliced expressions (meaning that they appear directly in the desugaring), or are they parsed in some other way peculiar to this literal form?
\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2, or might it capture an unseen binding of the same identifier used in the desugaring of Line 3?
\item \textbf{(Context Dependence)} If \li{w}, on Line 1, is renamed, could that possibly break Line 3, or change its meaning? In other words, might the desugaring of Line 3 assume that some variable identified as \li{w} is in scope (even though \li{w} is not mentioned in the text of Line 3)?
\item \textbf{(Typing)} What type does \li{y} have?
\end{enumerate}

% In short, the problem is that library providers cannot reason modularly about syntactic determinism (because other extensions might define overlapping forms), and if the desugaring of the program text is held abstract, programmers can no longer reason about types, binding and segmentation (i.e. answer questions like those above.) This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter a wide variety of libraries \cite{DBLP:conf/sac/LammelPS11}. 
Forcing the programmer to directly examine the desugaring of the program text, or the logic that produced that desugaring, in order to answer these sorts of basic questions  can defeat the purpose of syntactic sugar -- decreasing cognitive cost \cite{Green89}. Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala -- the binding discipline of the language ensures that programmers can reason lexically about binding and renaming, and types mediate abstraction over function and module implementations.  %From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.

This paper details a more reasonable approach: typed literal macros (TLMs). TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms}, which syntactically subsume other literal forms because their bodies are left initially unparsed according to the fixed context-free syntax of the language (i.e. they are initially parsed like raw string literals.) The applied TLM parses the literal body and generates an expansion during a phase known as typed expansion, which generalizes the usual typing phase. For example, Fig. \ref{fig:K-tsm-example} applies a TLM, \li{$kquery}, to a backtick-delimited generalized literal form to express the ``unreasonable'' example from Fig. \ref{fig:K-dialect}. TLMs come equipped with uniquely powerful abstract reasoning principles. We will characterize these reasoning principles absolutely precisely as we proceed, but for now, let us develop some informal intuitions -- the client programmer can reason about Line 3 as follows:
\begin{enumerate}
\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is solely responsible for parsing and expansion of the literal body. The context-free syntax of the language is fixed, so the TLM provider can modularly establish that the syntax the TLM implements is deterministic.
\item \textbf{(Segmentation)} The output that a TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced segments, which appear directly in the expansion, from segments that are parsed in some other way by the TLM. The segmentation can be communicated by a code editor using secondary notation, e.g. colors in this document. By examining the segmentation of Line 3, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions, whereas \li{R} and the other characters are parsed in some other way.
\item \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2. It cannot capture a coincidental binding of \li{x} in the expansion of Line 3.
\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope. The client is free to rename \li{w}. (As it turns out, enforcing context independence while also allowing macros to parse literal bodies arbitrarily is not straightforward. We return to this point in Sec. \ref{sec:existing-approaches}.)
\item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that the expansion must have. We will see an example of a TLM definition in the next section. Moreover, each spliced segment in the segmentation also comes paired with the type it is expected to have. This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
\end{enumerate}

@@

\begin{figure}
\vspace{-4px}
\begin{lstlisting}
  val w = compute_w()
  val x = compute_x(w)
  val y = $kquery `SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL`
\end{lstlisting}
\vspace{-10px}
\caption{TLMs make examples like the one from Fig. \ref{fig:K-dialect} more reasonable.}
\vspace{-12px}
\label{fig:K-tsm-example}
\end{figure}

The remainder of this document is organized as follows.
\begin{itemize}
\item Sec. \ref{sec:setsms} begins by defining a {simple expression TLM} for XHTML and describes initial parsing of generalized literal forms, typed expansion and expansion validation in more detail.
\item Sec. \ref{sec:sptsms} introduces {simple pattern TLMs}, i.e. TLMs that generate patterns. %The hygiene condition for pattern TLMs is interesting.
\item Sec. \ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above.

@@

% We next develop a more sophisticated account of TLMs in Chapters \ref{chap:ptsms} and \ref{chap:static-eval}, with the aim of working out details necessary to integrate TLMs into full-scale general-purpose languages like ML or Scala. %These two chapters constitute Part \ref{part:parametric-tsms} of our contributions.
\item Sec. \ref{sec:ptsms} adds type functions and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e. TLMs that take type and module parameters. Parameters serve two purposes:
\begin{enumerate}
\item They allow a single TLM to operate not just at a single type, but over a type- and module-parameterized family of types. For example, we consider a parametric TLM for dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the range type.
\item Parameters also give expansions access to terms other than those explicitly spliced in. %This is particularly convenient for working with abstract types \emph{a la} ML. 
Partial parameter application is possible, which lowers the syntactic cost of the strict context independence discipline that the system enforces.
\end{enumerate}
This section culminates with a calculus that extends the simpler calculus of Sec. \ref{sec:setsms-formally} with type functions, ML-style modules and parametric TLMs.
\item Sec. \ref{sec:static-eval} considers the topic of term evaluation during the typed expansion phase, i.e. \emph{static evaluation}, in  more detail, and also gives examples of TLMs that are useful for defining other TLMs, e.g. TLMs that implement parser generators and quasiquotation.
\item Sec. \ref{sec:existing-approaches} compares TLMs to the related work, including closely related work by \citet{TSLs,sac15} on type-specific languages, and the long history of work on macro systems.
\item Sec. \ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a brief discussion of present limitations and future research directions.
\end{itemize}

Due to space limitations, many standard or straightforward technical details of the calculi defined in Sec. \ref{sec:setsms-formally} through Sec. \ref{sec:static-eval} are relegated to the supplemental material. The examples that we give are written in an emerging dialect of ML called VerseML that supports TLMs as described herein. Our design should straightforwardly generalize to other languages with a broadly similar philosophy regarding types and binding (e.g. Haskell, Scala and others.) Languages that have a disciplined binding structure but that lack a rich static type structure, e.g. Racket and other Lisp derivatives, may also benefit from TLMs -- one can apply the results of this paper by deploying the usual type theorist's trick of casting such a language as a ``{unityped}'' language \cite{pfpl,scott1980lambda}. % For the present purposes, we can consider VerseML as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs. %VerseML is, as its name suggests, a conceptual descendent of ML. It diverges from other dialects of ML that have a similar type structure in that it has a bidirectional type system \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanism of TLM implicits described in Chapters \ref{chap:tsls} and \ref{chap:ptsms}. 
%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of VerseML in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TLMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work. 
% We will not formally define these features mainly to avoid unnecessarily complicating our presentation with details that are not essential to the ideas presented herein. As such, 
% All examples written in VerseML should be understood to be informal motivating material for the subsequent formal material. %We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble. %The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification. 
%We will give a brief overview of these languages are organized in Sec. \ref{sec:VerseML}.

@@

\end{lstlisting}
\vspace{-5px}
\caption[Definitions of various types used by TLM definitions.]{Definitions of various types used by TLM definitions. We assume that these are available ambiently.}
\label{fig:indexrange-and-parseresult}
\label{fig:candidate-exp-verseml}
\vspace{-5px}
\end{figure}
